---
title: 'MECH481A6: Engineering Data Analysis in R'
subtitle: 'Chapter 4 Homework: Visualizing Ozone Data' 
author: 'Michael Thill'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

*Note*: If you haven't installed $LaTex$, change the output mode in the above
YAML to `html_document` for ease of knitting and homework submission. 

This R Markdown (.Rmd) file is a template for your Chapter 4 Homework. Do 
everything within this file. Make it your own, but be careful not to change
the code-figure-text integration I set up with the code appendix and the 
global options. If you have used R Markdown before and are comfortable with the
extra options, feel free to customize to your heart's desire. In the end,
we will grade the **knitted** PDF or HTML document from within your private
GitHub repository. Remember to make regular, small commits (e.g., at least one
commit per question) to save your work. We will grade the latest knit, as long
as it occurs *before* the start of the class in which we advance to the next
chapter. As always, reach out with questions via GitHub Issues or during
office hours.

```{r global-options, include = FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = "../figs/",
                      echo = FALSE, warning = FALSE, message = FALSE)
```

# Ozone data

The corresponding data file (.csv) contains *hourly* ozone data from two sites
in Fort Collins. You should already have this file in your `/data` folder.

# Preparation

You completed the following steps in your Chapter 3 Homework. If correct, you
should copy-paste the code into this R Markdown document; FYI, you cannot 
`source()` R Markdown files for use of its output in another file because they
are intended to be self-contained and reproducible. Therefore, you need to 
copy-paste parts of your Chapter 3 Homework into this document and adjust 
the pathnames, if needed.

## Load R packages

```{r load-packages}
# load packages for current R session

library("tidyverse")

```

## Import, select, and clean data

Recreate the pipe of `dplyr` functions that you used to import the data, select
and rename the variables listed below, drop missing observations, and assign
the output as a `tibble` (*not in that particular order*).

- `sample_measurement` renamed as `ozone_ppm` (ozone measurement in ppm)
- `datetime` (date in YYYY-MM-DD format and time of measurement in HH:MM:SS)

```{r import-data}
# ozone: import, select, drop missing observations, rename

ozone_data <- read_csv("../Data/ftc_o3.csv") %>%
  select(sample_measurement, datetime) %>%
  drop_na() %>%
  rename(ozone_ppm = sample_measurement)

```

## Examine Data

Examine the structure and contents of the dataframe to confirm the file
imported and was manipulated properly. 

```{r missing-data}
# examine dataframe object 

head(ozone_data)

```

# Question 1: `ggplot2` time series

Using `ggplot` and the corresponding `geom`, create a time series of ozone
measurement across time. Warning: This plot will have a very poor
ink-to-information ratio. Ugly plots are okay when you are just exploring data.

```{r ozone-time}
# create basic time series using ggplot2 package

ggplot(ozone_data, aes(x = datetime, y = ozone_ppm)) +
  geom_line()

```

# Question 2: Base R equivalent

For comparison, what function could you use to create a time series of these
data in base R? How does the syntax of this function compare to that of
`ggplot()`?

# Question 3: `ggplot` object

Excluding the geom, assign the plot from Question 1 as a `ggplot` object with 
a descriptive name.

```{r ozone-object}
# create base layer of ozone time series (no geom) and save to object
```

# Question 4: `geom`

Now, `geom_point()` to the `ggplot` object using the following syntax:
`object_name + geom_point()`. Remember, you have already defined the `aes()`
in the `ggplot` object in Question 3.

```{r ozone-geom}
# add layer to ggplot object
```

# Question 5: `ggplot` layers

Call and examine object within the R Markdown, Console, or using View. How
many layers does this `ggplot` object contain? Why?

# Question 6: theme

Next, add the `ggplot2` theme  of your choice to the `ggplot` object with
`theme_*()` function prefix.

```{r ozone-theme}
# add theme to ggplot object
```

# Question 7: additions

In addition to assigning a plot as a `ggplot` object, one can also assign 
aspects of the figure such as axis labels and titles to an object for later 
use. For example: 

```{r ozone-labels, eval=FALSE}
# create new object with ggplot labels 
ozone_labels <- labs(x = "Time of Measurement (YYYY-MM-DD HH:MM:SS)",
                     y = "Ozone Concentration (ppm)",
                     title = "Hourly Ozone Measurements in Fort Collins, CO")
```

Using this technique and the same additive approach (`ggplot_object + ... + title`) from Questions 5 and 6, add a title and revise the axis labels.

```{r ozone-time-label}
# add title and axis labels to time series
```

# Question 7: ways to see more granularity  

The time series from the previous questions does not look nice. It is hard
to discern granular patterns in the data because of their sheer density; there
are too many hourly measurements over time. We could look at the data on 
different time scales, but, because we have not discussed how to manipulate
dates and times, we will instead focus first on adding transparency to the data 
points using the `alpha = ` aesthetic.  Try recreating the time-series plot with 
`alpha = 0.2`.

```{r ozone-time-label-alpha}
# add alpha aesthetic to the geom_point()

```

Alternatively, we could examine just the ozone measurements that
exceed the threshold (0.070 ppm) set by the Environmental Protection Agency.
Filter the dataset to the ozone levels exceeding 0.070 ppm, and use these data
to construct a time series plot with time of measurement on the x-axis and 
ozone concentration measurement on the y-axis. Remember to add the relevant
`geom`, title, subtitle, axis labels, and your choice of theme. 

```{r high-ozone-plot}
# filter data to ozone concentration measurements exceeding 0.070 ppm 

# time series of high ozone measurements
```

# Question 8: seasonality

Based on the time series, do you see any seasonal pattern for higher levels
of ozone? Describe what you see.

# Question 9: proportion

What proportion of ozone measurements exceed the EPA guidelines? Instead of
plugging in the actual values, make R figure out the length of each vector and
the corresponding proportion in one line of code.

```{r ozone-perc}
# calculate proportion of ozone measurements that exceed 0.070 ppm 
```

# Question 10: `ggplot2` extensions

Navigate to [this website](https://exts.ggplot2.tidyverse.org/gallery/) and
browse the `ggplot2` extensions. These themes can be very useful, so it's good
to be aware of them. Which theme would be appropriate for your own research or
senior project, and why? How would you use it? Briefly describe your data
and why the extension would improve the data visualization and communication.

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
